Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-30 paper2.qmd
section:theoretical; process:write

::: {#fig-caseV_issues layout-ncol=2}

@@

<!-- 1. The SCMs -->
Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs. SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables.

For instance, @fig-CJ1_scm display the SCM relating the CJ outcome $(O_{kjabi_{1}i_{2}})$, discriminal difference $(D_{kjabi_{1}i_{2}})$, perceived discriminal processes for each text within each student $(PT_{kjai_{1}}, PT_{kjbi_{2}})$, the "true" discriminal processes for each text within each student $(T_{kjai_{1}}, T_{kjbi_{2}})$, and the judges' biases $(B_{kj})$. Notice that the structural equations use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

::: {#fig-CJ1 layout-nrow=2}

::: {#fig-CJ1_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{kjabi_{1}i_{2}} & := f_{O}(D_{kjabi_{1}i_{2}}) \\
  D_{kjabi_{1}i_{2}} & := f_{D}(PT_{kjai_{1}}, PT_{kjbi_{2}}) \\
  PT_{kjai_{1}} & := f_{PT}(T_{ai_{1}}, T_{bi_{2}}, B_{kj})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_04.png){#fig-CJ1_DAG fig-align="center" fig-pos="H" width=45%}

@@

@@

<!-- ![](/images/png/CJ_TM_01.png){#fig-CJ_TM_01 width=23%} -->

<!-- ![](/images/png/CJ_TM_02.png){#fig-CJ_TM_02 width=32%} -->

<!-- ![](/images/png/CJ_TM_03.png){#fig-CJ_TM_03 width=45%} -->

without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in

![](/images/png/CJ_TM_05.png){#fig-CJ_TM_05 width=44%}

<!-- ![](/images/png/CJ_TM_06.png){#fig-CJ_TM_06 width=59%} -->

<!-- ![](/images/png/CJ_TM_07.png){#fig-CJ_TM_07 width=68%} -->

<!-- ![](/images/png/CJ_TM_08.png){#fig-CJ_TM_08 width=68%} -->

<!-- ![](/images/png/CJ_TM_09.png){#fig-CJ_TM_09 width=68%} -->

![](/images/png/CJ_TM_10.png){#fig-CJ_TM_10 width=68%}

@@

<!-- ![](/images/png/CJ_TM_11.png){#fig-CJ_TM_11 width=80%} -->

![](/images/png/CJ_TM_12.png){#fig-CJ_TM_12 width=80%}

@@

<!-- ![](/images/png/CJ_TM_13.png){#fig-CJ_TM_13 width=90%} -->

![](/images/png/CJ_TM_14.png){#fig-CJ_TM_14 width=90%}

![](/images/png/CJ_TM_15.png){#fig-CJ_TM_15 width=90%}

>>>>

::: {#fig-casev_issues layout-ncol=2}

@@

<!-- 1. The SCMs -->
Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs. SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

For instance, following the scenario outline in @sec-theoretical, @fig-cj04_scm displays the SCM that relates the outcome of the CJ experiment $(O_{abhijk})$, to several endogenous variables $V$. These variables include the discriminal difference $(D_{abhijk})$, perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$, formally expressed as $V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$. The structural equations $\{f_{O},f_{D},f_{PT}\}$ define the dependencies among these variables. Notably, in this scenario, the judges produce a dichotomous outcome $(O_{abhijk}) = \{0,1\}$. Moreover, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ represents the comparison number. Regarding the $k$ index, the index accounts for cases in which judges perform multiple comparisons of the same texts written by the same students, as the one discussed in @sec-theoretical.

@@

<!-- ![](/images/png/CJ_TM_01.png){#fig-cj01_dag fig-align="center" fig-pos="H" width=23%} -->

<!-- ![](/images/png/CJ_TM_02.png){#fig-cj02_dag fig-align="center" fig-pos="H" width=32%} -->

<!-- ![](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=45%} -->

::: {#fig-cj04 layout-nrow=2}

::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=45%}

@@

Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in

@@

::: {#fig-cj05 layout-nrow=2}

::: {#fig-cj05_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
\end{aligned}
$$

![DAG](/images/png/CJ_TM_05.png){#fig-cj05_dag fig-align="center" fig-pos="H" width=44%}

<!-- ![](/images/png/CJ_TM_06.png){#fig-cj06_dag fig-align="center" fig-pos="H" width=59%} -->

<!-- ![](/images/png/CJ_TM_07.png){#fig-cj07_dag fig-align="center" fig-pos="H" width=68%} -->

<!-- ![](/images/png/CJ_TM_08.png){#fig-cj08_dag fig-align="center" fig-pos="H" width=68%} -->

<!-- ![](/images/png/CJ_TM_09.png){#fig-cj09_dag fig-align="center" fig-pos="H" width=68%} -->

::: {#fig-cj10 layout-nrow=2}

::: {#fig-cj10_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

![DAG](/images/png/CJ_TM_10.png){#fig-cj10_dag fig-align="center" fig-pos="H" width=68%}

@@

<!-- ![](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=80%} -->

::: {#fig-cj12 layout-nrow=2}

::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J})
\end{aligned}
$$

![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=80%}

@@

<!-- ![](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=90%} -->

::: {#fig-cj14 layout-nrow=2}

::: {#fig-cj14_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

![DAG](/images/png/CJ_TM_14.png){#fig-cj14_dag fig-align="center" fig-pos="H" width=90%}

::: {#fig-cj15 layout-nrow=2}

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

![DAG](/images/png/CJ_TM_15.png){#fig-cj15_dag fig-align="center" fig-pos="H" width=90%}
